Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-16 cordillera-cycle.tex
Finish to reference the introduction.

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera were more extensive than today. At the last glacial maximum, a
contiguous blanket of ice extended from the Alaska Range in the north to the
North Cascades in the south (Fig.~\ref{fig:locmap}). This ice mass is known as
the former Cordilleran ice sheet \citep{Dawson.1888}.

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle \citep{Clague.1989}. However, former ice thickness and the ice
sheet's contribution to the last glacial maximum sea-level low-stand remain
poorly constrained. Moreover, our understanding of the glaciation history is
mainly restricted to the deglaciation phase and a regional scale, while little
is known about the ice sheet evolution prior to the last glacial maximum extent
\citep{Kleman.etal.2010}. Although time-evolving, whole ice-sheet
reconstructions of glacial advance and retreat patterns are available for the
Eurasian and Laurentide and ice sheets \citep{Kleman.etal.1997,
Kleman.etal.2010}, such is not the case for the Cordilleran ice sheet, where
complex arrangements of ice flow directions emanating from multiple glaciation
centres have been identified \citep{Prest.etal.1968, Clague.1989}, but remain
poorly understood \citep{Kleman.etal.2010}.

The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against
field-based evidence to perform a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial cycle. Although
numerical modelling has been established as a useful tool to improve our
understanding of the Cordilleran ice sheet more than twenty years ago
\citep{Jackson.Clague.1991,Robert.1991}, the ubiquitously
mountainous topography of the region has presented a major challenge to its
application, which can only be overcome by recent development of numerical ice
sheet models and underlying scientific computing tools \aref.
In addition to the topographic
complexity, strong climatic contrasts characteristic of the North American
Cordillera require the use of a high-resolution temperature and precipitation
forcing as input data to the ice sheet model \citep{Seguinot.etal.2013}.

Because past climate conditions are subject to considerable uncertainty, our
palaeoclimate forcing is a simplistic approximation of time-evolved temperature
and precipitation fields derived from a combination of a present-day
atmospheric reanalysis \citep{Mesinger.etal.2006}, lapse-rate corrections,
and temperature offset time series. The latter are
obtained by scaling six different palaeo-temperature reconstructions from
proxy records around the globe, including two \chem{\delta^{18}O} records from
Greenland ice cores \aref, two \chem{\delta^{18}O} records from Antarctic ice cores \aref,
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores \aref. Model output is compared to geomorphological evidence in terms
of timing and extent of glaciation and patterns of deglaciation.

>>>>

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera were more extensive than today. At the last glacial maximum, a
contiguous blanket of ice extended from the Alaska Range in the north to the
North Cascades in the south \citep[Fig.~\ref{fig:locmap};][Plate XIV]
{Chamberlin.1894}. This ice mass is known as the former Cordilleran ice sheet
\citep{Dawson.1888}.

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle \citep{Clague.1989, Dyke.2004}. However, former ice thickness
and the ice sheet's contribution to the last glacial maximum sea-level
low-stand remain poorly constrained. Moreover, our understanding of the
glaciation history is mainly restricted to the deglaciation phase and a
regional scale, while little is known about the ice sheet evolution prior to
the last glacial maximum extent \citep{Kleman.etal.2010}. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Eurasian and Laurentide and ice sheets
\citep{Kleman.etal.1997, Kleman.etal.2010}, such is not the case for the
Cordilleran ice sheet, where complex arrangements of ice flow directions
emanating from multiple glaciation centres have been identified
\citep{Prest.etal.1968, Clague.1989}, but remain poorly understood
\citep{Kleman.etal.2010}.

The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle. Although numerical modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet more than twenty years
ago \citep{Jackson.Clague.1991,Robert.1991}, the ubiquitously mountainous
topography of the region has presented a major challenge to its application.
In fact, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep[e.g.,][]{Bueler.Brown.2009} allows
high-resolution numerical modelling of palaeo-glacier flow on mountainous
terrain over millenial time scales \citep{Golledge.etal.2012}. In the North
American Cordillera, the topographic complexity induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as input to the ice sheet model
\citep{Seguinot.etal.2013}.

Because past climate conditions are subject to considerable uncertainty, our
palaeoclimate forcing is a simplistic approximation of time-evolved temperature
and precipitation fields derived from a combination of a present-day
atmospheric reanalysis \citep{Mesinger.etal.2006}, lapse-rate corrections,
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two \chem{\delta^{18}O} records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two \chem{\delta^{18}O}
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores \citep{Herbert.2001}. Model output is compared to
geomorphological evidence in terms of timing and extent of glaciation and
patterns of deglaciation.
